A Spectral Analysis of the Domain-Decomposed Adjoint Neumann-Ulam Method

### 2012-11-20 doc/mandc2013/notes/spectral_theory.tex
adding makefile and stencil fig to spectral theory doc
updates to spectral note

\date{\today}
\title{A Spectral Theory for the Domain-Decomposed Adjoint
  Neumann-Ulam Method}
\begin{document}
\maketitle

@@

  The serial and parallel behavior of the adjoint Neumann-Ulam method
  for solving linear operator equations is analyzed using the spectral
  properties of the linear operator. Relationships for the average
  length of the Markov chains and the fraction of histories leaking
  from a parallel domain are made with respect to the eigenvalues of
  the linear operator. The one-speed, two-dimensional neutron
  diffusion equation is used as a model problem to test the spectral
  theory for symmetric operators. These relationships will serve as
  guidelines for selecting an appropriate parallel decomposition
  stratagy and provide a basis for performance models for future
  implementations of a parallel adjoint Neumann-Ulam method.

@@

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} \cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6\Delta^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}

We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6\Delta^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}] + \Sigma_a \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{S}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{S}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unkown fluxes.

To close the system a non-reentrant current condition is applied
to all boundaries of the domain.

@@

%%---------------------------------------------------------------------------%%
\section{Length of Random Walks}

>>>>

\date{\today}
\title{A Spectral Analysis of the Domain-Decomposed Adjoint
  Neumann-Ulam Method}
\begin{document}
\maketitle

@@

  The serial and parallel behavior of the adjoint Neumann-Ulam method
  for solving linear operator equations is analyzed using the spectral
  properties of the linear operator. Relationships for the average
  length of the random walks and the fraction of histories leaking
  from a parallel domain are made with respect to the eigenvalues of
  the linear operator. The one-speed, two-dimensional neutron
  diffusion equation is used as a model problem to test the spectral
  theory for symmetric operators. These relationships will serve as
  guidelines for selecting an appropriate parallel decomposition
  stratagy and provide a basis for performance models for future
  implementations of a parallel adjoint Neumann-Ulam method.

@@

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} \cite{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6\Delta^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.25}{\input{stencil.pdftex_t}}
  \end{center}
  \caption{\textbf{Nine-point Laplacian stencil.}}
  \label{fig:stencil}
\end{figure}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6\Delta^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}] + \Sigma_a \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{S}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{S}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unkown fluxes. We
note here that the diffusion operator is symmetric.

To close the system a non-reentrant current condition is applied to
all boundaries of the domain. For our problems, we choose the
formulation of Duderstadt by assuming the flux is zero at some ghost
point beyond the grid. Consider for example the equations on the $i=0$
boundary of the domain:
\begin{multline}
  -\frac{1}{6\Delta^2}[4 \phi_{-1,j} + 4 \phi_{1,j}
    + 4 \phi_{0,j-1} + 4 \phi_{0,j+1} + \phi_{-1,j-1}\\ +
    \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20
    \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
  \label{eq:x_min_bnd}
\end{multline}
Here we note some terms where $i=-1$ and therefore are representative
of grid points beyond the boundary of the grid. We set the flux at
these points to be zero, giving a valid set of equations for the $i=0$
boundary:
\begin{multline}
  -\frac{1}{6\Delta^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
  \label{eq:x_min_bnd}
\end{multline}

@@

%%---------------------------------------------------------------------------%%
\section{Random Walk Lengths}
